---
title: "Strava Hiking Regression"
author: "Luke Beebe"
date: "2023-12-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gpx)
library(tidyverse)
library(rstanarm)
library(ggplot2)
library(bayesplot)
library(tidybayes)
library(broom.mixed)
```

## The project

My final project for Intro to Bayesian Analysis explores my personal data of 9 hikes from 4 states over the end of the summer I logged to and downloaded from Strava. My hope is to find a relationship between my HR, the elevation I've gained given the hike, and my speed. My hope is to use this as a way to set thresholds going forward. The data was exported as 8 seperate .gpx files that I preprocessed into one file.

```{r gpx}
setwd("C:/Users/bbkid/Documents/School/Fall 2023/Bayesian/Final Project")
hiking_data <- read_gpx("hiking data.gpx")$tracks
```

Let's see what these dataframes look like.

```{r data1}
head(hiking_data$`Slaughter trail`)
```
My next task is to calculate the change in elevation, latitude, longitude, and copy the heart rate from second to second (row to row).

In increments of 'time' c(5, 10, 15, 20, 30, 45, 60) minutes I summed the total latitude, longitude, and elevation differences and converted the values to ele_ft (elevation gain, loss in feet per 'time' min) and mph (miles per hour per 'time' min). I also saved my heart rate at each second calculated its average, saving to HR_avg. I combined these values into a dataframe df with i (the hike number). 

I chose these separate times to compare their correlations with our response variable. I did not know which increments I should lump these measurements into, but after running this search I found that 20 min seemed like the best value to work with the highest average correlation.

I commented out the code that runs the search to create the dataframe we'll use for the rest of this project with the highest correlated variables (pooled).

```{r data2}
#for(time in c(5, 10, 15, 20, 30, 45, 60)){
for(time in c(20)){
  ele_deltas <- lat_deltas <- long_deltas <- df <- NULL
  for(i in 1:length(hiking_data)){
    hike <- hiking_data[[i]]
    h <- as.character(i)
    HR_sum <- 0
    for(j in 1:(length(hike$Elevation)-1)){
      ele_delta <- hike$Elevation[j+1] - hike$Elevation[j]
      lat_delta <- hike$Latitude[j+1] - hike$Latitude[j]
      long_delta <- hike$Longitude[j+1] - hike$Longitude[j]
      HR <- as.integer(hike$hr[j+1])
      ele_deltas <- append(ele_deltas, ele_delta)
      lat_deltas <- append(lat_deltas, lat_delta)
      long_deltas <- append(long_deltas, long_delta)
      HR_sum <- HR_sum + HR
      if(j%%(60*time)==0){
        ele_gain_ft <- (3.28084)*sum(ele_deltas[ele_deltas>0])
        ele_loss_ft <- abs((3.28084)*sum(ele_deltas[ele_deltas<0]))
        mph <- (60/time)*(3280.84)*(10000/90)*sqrt(sum(lat_deltas)^2+sum(long_deltas)^2)/5280
        HR_avg <- HR_sum/(60*time)
        if(j==time && i==1){
          df <- data.frame(ele_gain_ft, ele_loss_ft, HR_avg, mph, h)
        }
        else{
          df <- rbind(df, data.frame(ele_gain_ft, ele_loss_ft, HR_avg, mph, h))
        }
        ele_deltas <- lat_deltas <- long_deltas <- NULL
        HR_sum <- 0
      }
    }
  }
  print(time)
  gain_r <- cor(df$ele_gain_ft, df$mph)
  loss_r <- cor(df$ele_loss_ft, df$mph)
  hr_r <- cor(df$HR_avg, df$mph)
  total_r_avg <- (abs(gain_r)+abs(hr_r)+abs(loss_r))/3
  print(paste("gain, mph:", gain_r))
  print(paste("loss, mph:", loss_r))
  print(paste("hr, mph:", hr_r))
  print(paste("avg:", total_r_avg))
}
```
I wanted to see what the df looked like before using it in ggplot, much nicer!
```{r data3}
df
```
```{r data4}
ggplot(df, aes(x=ele_gain_ft, y=mph, group=h, color=h)) +
  geom_point(color="black", size=3) +
  geom_smooth(size=1.5, method = "lm", se = FALSE) +
  scale_color_brewer(palette = "Spectral")
ggplot(df, aes(x=ele_loss_ft, y=mph, group=h, color=h)) +
  geom_point(color="black",size=3) +
  geom_smooth(size=1.5, method = "lm", se = FALSE) +
  scale_color_brewer(palette = "Spectral")
ggplot(df, aes(x=HR_avg, y=mph, group=h, color=h)) +
  geom_point(color="black",size=3) +
  geom_smooth(size=1.5, method = "lm", se = FALSE) +
  scale_color_brewer(palette = "Spectral")

ggplot(df, aes(x=ele_gain_ft, y=mph, group=h, color=h)) +
  geom_point(color="black",size=3) +
  geom_smooth(size=1.5, method = "lm", se = FALSE) +
  facet_wrap(~ h) + scale_color_brewer(palette = "Spectral")
ggplot(df, aes(x=ele_loss_ft, y=mph, group=h, color=h)) +
  geom_point(color="black",size=3) +
  geom_smooth(size=1.5, method = "lm", se = FALSE) +
  facet_wrap(~ h) + scale_color_brewer(palette = "Spectral")
ggplot(df, aes(x=HR_avg, y=mph, group=h, color=h)) +
  geom_point(color="black",size=3) +
  geom_smooth(size=1.5, method = "lm", se = FALSE) +
  facet_wrap(~ h) + scale_color_brewer(palette = "Spectral")
```

Right away, we see that ele_gain_ft and ele_loss_ft are similar, as we know if we're returning to the same starting point, then the gain and loss will be the same. I would cut it, but it seems like the more elevation I lose, my speed increases more than elevation gain. And, because this is for 20 min intervals, it may tell us more if I'm mostly gaining or losing during one interval. HR_avg is also trending up, as if I exert more energy, I go faster. This is to be assumed in a 20 min time period, but not sustainable any longer, which is why I should put a disclaimer that these predictions should only be made for sustainable heart rates below my lactate threshold.

I chose to compare the different hikes, because the one thing my data doesn't account for is the grade or type of hike. For example, Mount Sneffels via the SW Ridge is a Colorado 14er class 3 scramble without many trail markers. I am climbing rock, and going much slower than what I would on Lower Blue Lakes, the trail I took to the campsite I slept at the night before. Mt. Sneffels is '3', and Lower Blue lake is '5' on the graphs above. You can see the lower speeds and higher heart rate across the board. You can also see Crags Trail, '1', as one of my faster hikes. That was in a different region of Colorado, after Sneffels, which was a smooth and gradual trail up, where I pushed it by myself.

Let's try a standard Normal regression model with pooled data, and another with an interaction term

```{r data5}
hiking_pooled <- stan_glm(mph ~ ele_gain_ft + ele_loss_ft + HR_avg + as.integer(h),
                        data = df, family = gaussian,
                        prior_intercept = normal(2,1),
                        prior = normal(0, 0.5, autoscale = T),
                        prior_aux = exponential(1, autoscale = T),
                        chains = 4, iter = 5000*2, seed = 12345)

hiking_hierarchal <- stan_glmer(mph ~ ele_gain_ft + ele_loss_ft + HR_avg + (HR_avg | as.integer(h)),
                        data = df, family = gaussian,
                        prior_intercept = normal(2,1),
                        prior = normal(0, 0.5, autoscale = T),
                        prior_aux = exponential(1, autoscale = T),
                        chains = 4, iter = 5000*2, seed = 12345)

```

Let's check some numerical diagnostics.

```{r linreg}
mcmc_trace(hiking_pooled, size=0.1)
mcmc_trace(hiking_hierarchal, size=0.1)

print("rhat")
rhat(hiking_pooled)
rhat(hiking_hierarchal)
print("neff ratio")
neff_ratio(hiking_pooled)
neff_ratio(hiking_hierarchal)

tidy(hiking_pooled, effects=c("fixed", "aux"),
     conf.int=T, conf.level=0.9)
tidy(hiking_hierarchal, effects=c("fixed", "aux"),
     conf.int=T, conf.level=0.9)

pp_check(hiking_pooled)
pp_check(hiking_hierarchal)


```

These look good!
